HO unification from object language to meta language

### 2024-03-28 paper/main.tex
tex updates

\begin{invariant}
  Any term in the substitution of \Ho is in normal form.
  TODO: move this to the right place
  TODO: definire normal form, maybe in definition 2.7
  \label{inv:subst-nf}
\end{invariant}
\begin{invariant}

  Any variable in the substitution with arity \elpiIn{N} is always fully applied
  TODO: move this to the right place
  TODO: definire normal form, maybe in definition 2.7
  \label{inv:fully-applied-vars}
\end{invariant}

@@

\begin{invariant}[Unification variable arity]\label{inv:uvaarity}
Each variable \elpiIn{A}
in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
\elpiIn{(uva A L)} is such that \elpiIn{(len L N)} holds
\label{invariant:arity}
\end{invariant}

@@

@@

\subsection{Limitations of by this basic scheme}

@@

Compiling term with \maybeeta terms forces us to determine if, $\lambda x. T_x$,
for any term $T$ having $x$ in scope, can be a $\eta$-expansion, i.e. under a
given substitution $\sigma$, we have $\sigma (\lambda x.T_x) = t$. This
\maybeeta detection is not a trivial operation as it may seems.
%
\begin{gather}
  \lambda x. f \appsep A_x\\
  \lambda x. f \appsep x \appsep A_x\\
  \lambda x. \lambda y. f \appsep A_x \appsep B_{y x}\\ \label{eq:lamlam}
\end{gather}

@@

A second way to progress \linketa, that we call \linketa deduplication, is when
\linkStore contains two \linketa $l_1$ and $l_2$ with a \lhs having the same
variable address. Let the \lhs of $l_1$ be \elpiIn{uva U R} and the \lhs of
$l_2$ be \elpiIn{uva V S}, then, by \cref{inv:fully-applied-vars}, \elpiIn{R}
and \elpiIn{S} have same scope. Let $t$ be the term obtained by replacing all
each name \elpiIn{S}$_i$ in the \rhs of $l_1$ with \elpiIn{R}$_i$, $t$ is
unified with the \rhs of $l_2$ and one of the two links between $l_1$ and $l_2$
is removed from \linkStore.

@@

We can note that the insertion of these rules for \elpiIn{progress1} do not
prevent the termination of \elpiIn{progress}, since: if a link is activated it
is removed from \linkStore and the recursive call to \elpiIn{progress} will have
a smaller list of links to recurse on. Moreover, link activation only runs
unification which are terminating instructions. If a link is deduplicated, the
termination of \elpiIn{progress} is still guaranteed since again we reduce
\linkStore and the instructions run by link deduplications are all terminating.
Finally, if a link is neither activated nor deduplicated, i.e. it remains
suspended, then the condition of the \elpiIn{if}-statement of \elpiIn{progress}
succeeds making it to terminate.

@@

\subsection{Example}

In this subsection we show two complete runs of the extended version of \Ue.

\paragraph{Example 1}
\begin{center}
  \elpiIn{flam x\ flam y\ fapp[fuva A, y, x] = flam x\ flam y\ x}
\end{center}
% $$\flambda xy.F \appsep y \appsep x = \lambda xy.x$$
\begin{itemize}
  \item After compilation: terms are $X0 = \lambda c0.\lambda c1.c0$ with links
        $\linketaM{c0}{X1_{c0}}{\lambda c1.X2_{c1~c0}}, \linketaM{}{X0}{\lambda
        c0.X1_{c0}}$ and mappings: $\mapping{C0}{X2}{2}$
  \item After unification $X0$ is set to $\lambda c0.\lambda c1.c0$
  \item This unification wakes up the second link, since its \lhs is now
        instantiated. $X0$ is unified with its \rhs, which assigns $X1_{c0} :=
        \lambda c1.c0$
  \item The instantiation of $X1$ wakes the first link, and $\lambda c1.c0$ is
        unified with $\lambda c1.X2_{c1~c0}$, setting $X2 _{c1~c2} := c2$ by
        pruning.
  \item All the links have been solved and the decompilation, assigns the \Fo{}
        variable \elpiIn{A} to $\lambda xy.y$ 
\end{itemize}

\paragraph{Example 2}
\begin{center}
  \elpiIn{flam x\ fapp[f, fapp[fuva F, x], x] = fapp[f, flam x\ x]}
\end{center}
% $$\flambda xy.F \appsep y \appsep x = \lambda xy.x$$
\begin{itemize}
  \item After compilation: terms are $X0 = f~\lambda c0.c0$ with links
        $\linketaM{}{X0}{\lambda c0.(f~X1_{c0}~c0)}$ and mappings:
        $\mapping{C0}{X1}{1}$
  \item After unification $X0$ is set to $f~\lambda c0.c0$
  \item This unification wakes up the link, since its \lhs is now
        instantiated. $X0$ is unified with its \rhs, which assigns $X1_{c0} :=
        \lambda c1.c0$
  \item The instantiation of $X1$ wakes the first link, and $\lambda c1.c0$ is
        unified with $\lambda c1.X2_{c1~c0}$, setting $X2 _{c1~c2} := c2$ by
        pruning.
  \item All the links have been solved and the decompilation, assigns the \Fo{}
        variable \elpiIn{A} to $\lambda xy.y$ 
\end{itemize}

\section{Enforcing Invariant~\ref{invariant:arity}}
\label{sec:invariant1}
Deduplicate mapping code etc...

>>>>

\begin{invariant}
  Any term in the substitution of \Ho is in normal form.
  TODO: move this to the right place
  TODO: definire normal form, maybe in definition 2.7
  \label{inv:subst-nf}
\end{invariant}

@@

\begin{invariant}[Unification variable arity]
  Each variable \elpiIn{A}
  in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
  \elpiIn{(uva A L)} is such that \elpiIn{(len L N)} holds
  \label{inv:uvaarity}
\end{invariant}

@@

\begin{invariant}

  TODO: dire che il mapping è bijective
  \label{inv:map-bijective}
\end{invariant}

@@

\subsection{Limitations of by this basic scheme}
\label{sec:basic-comp-limitations}

@@

Compiling term with \maybeeta terms forces us to determine if, $\lambda x. T_x$,
for any term $T$ having $x$ in scope, can be a $\eta$-expansion, i.e. under a
given substitution $\sigma$, we have $\sigma (\lambda x.T_x) = t$. This
\maybeeta detection is not a trivial operation as it may seems.
%
\begin{gather}
  \lambda x. f \appsep A_x\\
  \lambda x. f \appsep x \appsep A_x\\
  \lambda x. \lambda y. f \appsep A_x \appsep B_{y x} \label{eq:lamlam}
\end{gather}

@@

A second way to progress \linketa, that we call \linketa deduplication, is when
\linkStore contains two \linketa $l_1$ and $l_2$ with a \lhs having the same
variable address. Let the \lhs of $l_1$ be \elpiIn{uva U R} and the \lhs of
$l_2$ be \elpiIn{uva V S}, then, by \cref{inv:uvaarity}, \elpiIn{R}
and \elpiIn{S} have same scope. Let $t$ be the term obtained by replacing all
each name \elpiIn{S}$_i$ in the \rhs of $l_1$ with \elpiIn{R}$_i$, $t$ is
unified with the \rhs of $l_2$ and one of the two links between $l_1$ and $l_2$
is removed from \linkStore.

@@

We can note that the insertion of these rules for \elpiIn{progress1} do not
prevent the termination of \elpiIn{progress}, since: if a link is activated it
is removed from \linkStore and the recursive call to \elpiIn{progress} will have
a smaller list of links to recurse on. Moreover, link activation only runs
terminating instructions (such as unification). If a link is deduplicated, the
termination of \elpiIn{progress} is still guaranteed since again we reduce
\linkStore and the instructions run by link deduplications are all terminating.
Finally, if a link is neither activated nor deduplicated, i.e. it remains
suspended, then the condition of the \elpiIn{if}-statement of \elpiIn{progress}
succeeds making it to terminate.

@@

% In this subsection we show two complete runs of the extended version of \Ue.

% \paragraph{Example 1}
% \begin{center}
%   \elpiIn{flam x\ flam y\ fapp[fuva A, y, x] = flam x\ flam y\ x}
% \end{center}
% % $$\flambda xy.F \appsep y \appsep x = \lambda xy.x$$
% \begin{itemize}
%   \item After compilation: terms are $X0 = \lambda c0.\lambda c1.c0$ with links
%         $\linketaM{c0}{X1_{c0}}{\lambda c1.X2_{c1~c0}}, \linketaM{}{X0}{\lambda
%         c0.X1_{c0}}$ and mappings: $\mapping{C0}{X2}{2}$
%   \item After unification $X0$ is set to $\lambda c0.\lambda c1.c0$
%   \item This unification wakes up the second link, since its \lhs is now
%         instantiated. $X0$ is unified with its \rhs, which assigns $X1_{c0} :=
%         \lambda c1.c0$
%   \item The instantiation of $X1$ wakes the first link, and $\lambda c1.c0$ is
%         unified with $\lambda c1.X2_{c1~c0}$, setting $X2 _{c1~c2} := c2$ by
%         pruning.
%   \item All the links have been solved and the decompilation, assigns the \Fo{}
%         variable \elpiIn{A} to $\lambda xy.y$ 
% \end{itemize}

% \paragraph{Example 2}
% \begin{center}
%   \elpiIn{flam x\ fapp[f, fapp[fuva F, x], x] = fapp[f, flam x\ x]}
% \end{center}
% % $$\flambda xy.F \appsep y \appsep x = \lambda xy.x$$
% \begin{itemize}
%   \item After compilation: terms are $X0 = f~\lambda c0.c0$ with links
%         $\linketaM{}{X0}{\lambda c0.(f~X1_{c0}~c0)}$ and mappings:
%         $\mapping{C0}{X1}{1}$
%   \item After unification $X0$ is set to $f~\lambda c0.c0$
%   \item This unification wakes up the link, since its \lhs is now
%         instantiated. $X0$ is unified with its \rhs, which assigns $X1_{c0} :=
%         \lambda c1.c0$
%   \item The instantiation of $X1$ wakes the first link, and $\lambda c1.c0$ is
%         unified with $\lambda c1.X2_{c1~c0}$, setting $X2 _{c1~c2} := c2$ by
%         pruning.
%   \item All the links have been solved and the decompilation, assigns the \Fo{}
%         variable \elpiIn{A} to $\lambda xy.y$ 
% \end{itemize}

\section{Enforcing Invariant~\ref{invariant:arity}}
\label{sec:invariant1}

In \cref{sec:basic-comp-limitations}, we have given two unification problems to
be run one after the other. The compilation of each term would generate the
following result:

\begin{description}
  \item[Unif problems:] $X = \lambda x.\lambda y.x, Z = f\appsep(\lambda x.x)$
  \item[Mappings:] $\mapping{F}{G}{1}, \mapping{F}{H}{2}$  
  \item[Links:] $\linketaM{x}{Y_{x}}{\lambda y.H_{y~x}}$,\\
                $\linketaM{}{X}{\lambda x.Y_{x}}$,\\
                $\linketaM{}{Z}{\lambda x.f \appsep G_{x}~x}$
\end{description}

We see that the \elpiIn{maybe-eta} as detected $\lambda xy.F \appsep y \appsep
x$ and $\lambda x.f\appsep (F \appsep x) \appsep x$ and replaced them with
respectively the \Ho vars $X$ and $Z$. $X$ is linked with $\lambda x.Y_x$, $Y$
has arity $1$ and is $\eta$-linked with $\lambda y.H\appsep y \appsep x$ and $Z$
is linked to the term $\lambda x.f \appsep G_{x}~x$. However, we can remark that
this compilation breaks \cref{inv:map-bijective}, since

Deduplicate mapping code etc...
